Multi-Label Speech Emotion Recognition Using 2D Convolutional Neural Networks

### 2019-09-07 research/paper/paper_body.tex
Change paper template to ACM SAC 2020

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{dblfloatfix}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{pifont}
\usepackage{textcomp}
\usepackage{multirow}
\usepackage[table,xcdraw]{xcolor}
\usepackage[strings]{underscore}

\graphicspath{{figures/}}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Multi-Label Speech Emotion Recognition Using 2D Convolutional Neural Networks}

\author{\IEEEauthorblockN{Brian Pho, Thomas Truong, Svetlana Yanushkevich}
\IEEEauthorblockA{\textit{Biometric Technologies Laboratory, Department of Electrical and Computer Engineering}\\
University of Calgary, Canada\\
\{brian.pho, thomas.truong, syanshk\}@ucalgary.ca}}

\maketitle

\begin{abstract}
Current speech emotion recognition systems often overlook the multi-label data that comes with databases. In this paper, we address the problem of whether machine learning can be used to detect multiple emotions in speech. We created a combined database consisting of four speech emotion-labeled databases, and we used it to train a 2D convolutional neural network to determine if the model could recognize multiple emotions in a speech sample. The model was able to classify the samples with an accuracy of 57.64\%. This shows that it is possible to apply machine learning to the problem of multi-label speech emotion recognition and to achieve a reasonable accuracy.\footnote{Code for this paper is available on Github at: \url{https://github.com/Brian-Pho/RVST598_Speech-Emotion-Recognition}}
\end{abstract}

\begin{IEEEkeywords}
multi-label, speech, emotion, recognition, neural networks
\end{IEEEkeywords}

\section{Introduction}
The field of affective computing studies the development of systems that can recognize, interpret, process, and simulate human affects. A subfield of affective computing that we are interested in is speech emotion recognition (SER). SER is the problem of recognizing which emotions are present in speech and SER has been approached in two ways:
\begin{itemize}
	\item Feature engineering
	\item Machine learning (ML)
\end{itemize}

@@

\bgroup
\def\arraystretch{2}
\begin{table*}
	\resizebox{\textwidth}{!}{%
		\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|c|c|c|}
			\hline
			\multicolumn{1}{|c|}{} & \multicolumn{13}{c|}{\textbf{Emotions}} \\ \cline{2-14} 
			\multicolumn{1}{|c|}{\multirow{-2}{*}{\textbf{Databases}}} & \textbf{Neutral} & \textbf{Anger} & \textbf{Disgust} & \textbf{Fear} & \textbf{Happy} & \textbf{Sad} & \textbf{Surprise} & \textbf{Calm} & \textbf{Excitement} & \textbf{Frustration} & \textbf{Amused} & \textbf{Sleepy} & \textbf{Bored} \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			\textbf{IEMOCAP} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark \\ \hline
			\textbf{CREMA-D} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			\textbf{TESS} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
			\textbf{RAVDESS} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			\textbf{MSP-IMPROV} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
			\textbf{SAVEE} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			\textbf{Emo-DB} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\ \hline
			\textbf{EmoV-DB} & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark & \cmark & \xmark \\ \hline
		\end{tabular}%
	}
	\caption{Comparison of databases with their labeled emotions.}
	\label{dbEmoTable}
\end{table*}
\egroup

@@

@@

\begin{figure*}[t]
	\centering
	\hspace{6mm}
	\includegraphics[width=\textwidth]{combined_db.png} 
	\caption{Proportions of each database, emotion, and label types in the combined database.}
	\label{combinedDb}
\end{figure*}

@@

\begin{figure}[]
	\centering
	\hspace{6mm}
	\includegraphics[width=\linewidth]{high_level_dataflow_diagram.png}
	\caption{High level overview of the processing stages that a speech sample goes through.}
	\label{highLevelDataFlowDiagram}
\end{figure}

@@

\begin{figure*}[h!]
	\centering
	\hspace{6mm}
	\includegraphics[width=\textwidth]{neural_network_architecture.png}
	\caption{Architecture for the speech emotion recognition network including the input layer and eight processing layers.}
	\label{neuralNetworkArchitecture}
\end{figure*}

@@

\bgroup
\def\arraystretch{2}
\begin{table}[h!]
	\centering
	\caption{The model's hyperparameters.}
	\label{modelHyperparams}
	{%
		\begin{tabular}{|l|r|}
			\hline
			\textbf{Hyperparameter} & \textbf{Value} \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			Input Dimensions & 278w x 200h x 1d \\ \hline
			Optimization Algorithm & Adam \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			Loss Measure & Binary Cross-entropy \\ \hline
			Accuracy Metric & Categorical Cross-entropy \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			Activation Function & Sigmoid \\ \hline
		\end{tabular}%
	}
\end{table}
\egroup

\bgroup
\def\arraystretch{2}
\begin{table}[h!]
	\centering
	\caption{The training hyperparameters.}
	\label{trainHyperparams}
	{%
		\begin{tabular}{|l|r|}
			\hline
			\textbf{Hyperparameter} & \textbf{Value} \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			Epochs & 20 \\ \hline
			Batch Size & 32 \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			Training Set & 17,341 \\ \hline
			Validation Set & 2,167 \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			Testing Set & 2,167 \\ \hline
		\end{tabular}%
	}
\end{table}
\egroup

@@

\begin{figure*}[h!]
	\centering
	\hspace{6mm}
	\includegraphics[width=\textwidth]{confusion_matrix.png}
	\caption{Confusion matrices for each emotion.}
	\label{confusionMatrix}
\end{figure*}

\bgroup
\def\arraystretch{2}
\begin{table}[]
	\centering
	\caption{Comparison of SER accuracy in literature.}
	\label{litacccompare}
	\begin{tabular}{|l|r|r|}
		\hline
		\multicolumn{1}{|c|}{\textbf{Research Work}} & \multicolumn{1}{c|}{\textbf{Method}} & \multicolumn{1}{c|}{\textbf{Accuracy (\%)}} \\ \hline
		\rowcolor[HTML]{F6F8FA}
		Zhao et al. \cite{Zhao2019} & CNN + LSTM & 52.1 \\ \hline
		Our Work & CNN & 57.6 \\ \hline
		\rowcolor[HTML]{F6F8FA}
		Etienne et al. \cite{Etienne2018} & CNN + LSTM & 61.7 \\ \hline
		Zhang et al. \cite{Zhang2019} & CNN & 63.9 \\ \hline
		\rowcolor[HTML]{F6F8FA}
		Fayek et al. \cite{Fayek2017} & CNN & 64.8 \\ \hline
		Yenigalla et al. \cite{Yenigalla2018} & CNN & 73.9 \\ \hline
		\rowcolor[HTML]{F6F8FA}
		Badshah et al. \cite {Badshah2019} & CNN & 84.3 \\ \hline
	\end{tabular}
\end{table}
\egroup

@@

\section*{Acknowledgment}
This work was funded by the Program for Undergraduate Research Experience (PURE) award granted by the University of Calgary.

\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv,references}

\end{document}

>>>>

\section{Introduction}

The field of affective computing studies the development of systems that can recognize, interpret, process, and simulate human affects. A subfield of affective computing that we are interested in is speech emotion recognition (SER). SER is the problem of recognizing which emotions are present in speech and SER has been approached in two ways:
\begin{itemize}
	\item Feature engineering
	\item Machine learning (ML)
\end{itemize}

@@

@@

\begin{table*}
	\centering
	\caption{Comparison of databases with their labeled emotions.}
	\label{dbEmoTable}
	\begin{tabular}{@{}lccccccccccccc@{}}
		\toprule
		Databases & \multicolumn{13}{c}{Emotions} \\ 
		\cmidrule{2-14}
		& Neutral & Anger & Disgust & Fear & Happy & Sad & Surprise & Calm & Excitement & Frustration & Amused & Sleepy & Bored \\ 
		\midrule
		IEMOCAP & x & x & x & x & x & x & x &   & x & x &   &   &   \\
		CREMA-D & x & x & x & x & x & x &   &   &   &   &   &   &   \\
		TESS & x & x & x & x & x & x & x &   &   &   &   &   &   \\
		RAVDESS & x & x & x & x & x & x & x & x &   &   &   &   &   \\
		MSP-IMPROV & x & x & x & x & x & x & x &   &   &   &   &   &   \\
		SAVEE & x & x & x & x & x & x & x &   &   &   &   &   &   \\
		Emo-DB & x & x & x & x & x & x &   &   &   &   &   &   & x \\
		EmoV-DB & x & x & x &   &   &   &   &   &   &   & x & x &   \\
		\bottomrule
	\end{tabular}%
\end{table*}

@@

\begin{figure*}
	\centering
	\hspace{6mm}
	\includegraphics[width=\textwidth]{combined_db.png} 
	\caption{Proportions of each database, emotion, and label types in the combined database.}
	\label{combinedDb}
\end{figure*}

@@

\begin{figure}
	\centering
	\hspace{6mm}
	\includegraphics[width=\linewidth]{high_level_dataflow_diagram.png}
	\caption{High level overview of the processing stages that a speech sample goes through.}
	\label{highLevelDataFlowDiagram}
\end{figure}

@@

\begin{figure*}
	\centering
	\hspace{6mm}
	\includegraphics[width=\textwidth]{neural_network_architecture.png}
	\caption{Architecture for the speech emotion recognition network including the input layer and eight processing layers.}
	\label{neuralNetworkArchitecture}
\end{figure*}

@@

\begin{table}
	\centering
	\caption{The model's hyperparameters.}
	\label{modelHyperparams}
	\begin{tabular}{@{}lr@{}}
		\toprule
		Hyperparameter & Value \\
		\midrule
		Input Dimensions & 278w x 200h x 1d \\
		Optimization Algorithm & Adam \\
		Loss Measure & Binary Cross-entropy \\
		Accuracy Metric & Categorical Cross-entropy \\
		Activation Function & Sigmoid \\
		\bottomrule
	\end{tabular}
\end{table}

\begin{table}
	\centering
	\caption{The training hyperparameters.}
	\label{trainHyperparams}
	\begin{tabular}{@{}lr@{}}
		\toprule
		Hyperparameter & Value \\
		\midrule
		Epochs & 20 \\
		Batch Size & 32 \\
		Training Set & 17,341 \\
		Validation Set & 2,167 \\
		Testing Set & 2,167 \\
		\bottomrule
	\end{tabular}
\end{table}

@@

\begin{figure*}
	\centering
	\hspace{6mm}
	\includegraphics[width=\textwidth]{confusion_matrix.png}
	\caption{Confusion matrices for each emotion.}
	\label{confusionMatrix}
\end{figure*}

\begin{table}
	\centering
	\caption{Comparison of SER accuracy in literature.}
	\label{litacccompare}
	\begin{tabular}{@{}lrr@{}}
		\toprule
		Research Work & Method & Accuracy (\%) \\
		\midrule
		Zhao et al. \cite{Zhao2019} & CNN + LSTM & 52.1 \\
		Our Work & CNN & 57.6 \\
		Etienne et al. \cite{Etienne2018} & CNN + LSTM & 61.7 \\
		Zhang et al. \cite{Zhang2019} & CNN & 63.9 \\
		Fayek et al. \cite{Fayek2017} & CNN & 64.8 \\
		Yenigalla et al. \cite{Yenigalla2018} & CNN & 73.9 \\
		Badshah et al. \cite {Badshah2019} & CNN & 84.3 \\
		\bottomrule
	\end{tabular}
\end{table}

@@

\begin{acks}
This work was funded by the Program for Undergraduate Research Experience (PURE) award granted by the University of Calgary.
\end{acks}
